---
title: "Project 3"
author: "Kara Belknap & Cassio Monti"
date: "2022-10-29"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Project 3

## Monti - Introduction


## Getting Started

Before we can begin our analysis, we must load in the following packages:

```{r libraries}
library(tidyverse)

```




## Read in the Data

Using the datafile `OnlineNewsPopularity.csv`, we will read in the data and add a new column corresponding to the type of data channel from which the data was classified. The new variable will be called `dataChannel`. Note that there are some rows that are unclassified according to the six channels of interest and those are indicated by `other`. 

Once the data column is created, we can easily subset the data using the `filter` function to create a new dataset for each data channel.


```{r dataImport}

rawData <- read_csv("OnlineNewsPopularity.csv")

rawDataChannel <- rawData %>%
  mutate(dataChannel = ifelse(data_channel_is_lifestyle == 1, "lifestyle", 
                              ifelse(data_channel_is_entertainment == 1, "entertainment", 
                                     ifelse(data_channel_is_bus == 1, "bus", 
                                            ifelse(data_channel_is_socmed == 1, "socmed", 
                                                   ifelse(data_channel_is_tech == 1, "tech", 
                                                          ifelse(data_channel_is_world == 1, "world", "other"))))))) %>%
  select(-data_channel_is_lifestyle, -data_channel_is_entertainment, -data_channel_is_bus, -data_channel_is_socmed, 
         -data_channel_is_tech, -data_channel_is_world)


lifestyleData <- rawDataChannel %>%
  filter(dataChannel == "lifestyle")

entertainmentData <- rawDataChannel %>%
  filter(dataChannel == "entertainment")

busData <- rawDataChannel %>%
  filter(dataChannel == "bus")

socmedData <- rawDataChannel %>%
  filter(dataChannel == "socmed")

techData <- rawDataChannel %>%
  filter(dataChannel == "tech")

worldData <- rawDataChannel %>%
  filter(dataChannel == "world")

```


## Summarizations

### Monti - Summary Stats


### Belknap - Summary Stats


### Monti - Graphs (3)



### Belknap - Graphs (3)


## Modeling

### Data Split

### Belknap - Linear Regression Model Explanation


### Monti - Linear Regression Model


### Belknap - Linear Regression Model



### Monti - Ensemble Tree-based Model


### Belknap - Ensemble Tree-based Model





### Belknap - Random Forest Model & Explanation




### Monti - Boosted Tree Model & Explanation




## Comparison & Conclusion - Monti or Belknap (whoever doesn't do automation of R Markdown)








