---
title: "Project 3"
author: "Kara Belknap & Cassio Monti"
date: "2022-11-5"
output:
  github_document:
    html_preview: false
    toc: true
params:
  channel: "lifestyle"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, error = FALSE, message = FALSE)
```

# Report for *`r params$channel`* Channel

This report contains Exploratory Data Analysis (EDA) about this data channel and a modeling section applying three regression methods. 

## Introduction

The objective of this analysis is to provide a comprehensive overview about publication metrics and their relationship with the number of shares that those publications presented during the study period. These data have been collected from Mashable website, one of the largest news websites from which the content of all the `r params$channel` channel articles published in 2013 and 2014 was extracted. These data were originally collected analyzed By Fernandes et al. (2015) work, in which the authors performed classification task comparing several machine learning algorithms. In the present study, the subset of the data used by Fernandes et al.(2015) corresponding to the data channel `r params$channel` is used for regression purposes. The response variable is the number of `shares` that the papers presented after publication. In other words, we will try to predict the number of shares the papers will have before publication. To perform the regression, Random Forest, Boosting, and Multiple Linear Regression are used. More information about the methods will be provided in further sections.    
   
   Some metrics have been calculated based on the information obtained from Marshable website. For instance, the Latent Dirichlet Allocation (LDA) was applied to the data set to identify the 5 top relevant topics and then measure the closeness of the current article to such topic. There are 5 relevance of topic metrics according to LDA:   
   
   - `LDA_00`: Closeness to LDA topic 0   
   - `LDA_01`: Closeness to LDA topic 1   
   - `LDA_02`: Closeness to LDA topic 2   
   - `LDA_03`: Closeness to LDA topic 3   
   - `LDA_04`: Closeness to LDA topic 4   

   Additionally, some quality metrics related to the keywords have been calculated and will be used in this analysis. These metrics represent the average number of shares for publications with worst, best, and average keywords. The classification of keywords under these groups was made by the authors of the original paper. The keyword metrics are shown below.   
   
   - `kw_avg_min`: Worst keyword (avg. shares)   
   - `kw_avg_max`: Best keyword (avg. shares)   
   - `kw_avg_avg`: Avg. keyword (avg. shares)   

   Article content metrics were also used in this study. These are general metrics about the body of the publication that can influence the number of shares of that paper. The content summary metrics are shown below.   
   
   -`num_videos`: Number of videos   
   -`n_tokens_content`: Number of words in the content   
   -`n_non_stop_unique_tokens`: Rate of unique non-stop words in the content   
   -`num_hrefs`: Number of links   
   -`num_self_hrefs`: Number of links to other articles published by Mashable   
   -`average_token_length`: Average length of the words in the content   

  These data were collected during 2013 and 2014 on daily basis. To represent time dependent information, a binary variable indicating whether the publication was made in a weekend or weekday, `is_weekend` is used.

## Required Packages

Before we can begin our analysis, we must load in the following packages:

```{r packages}

library(tidyverse)
library(caret)
library(PerformanceAnalytics)
library(knitr)

```

`Tidyverse` is used for data management and plotting through dplyr and ggplot packages. `Caret` package is used for data splitting and modeling. `knitr` package is used to provide nice looking tables. `PerformanceAnalytics` is used for nice correlation plots assisting in the visualization.

## Read in the Data

Using the data file `OnlineNewsPopularity.csv`, we will read in the data and add a new column corresponding to the type of data channel from which the data was classified. The new variable will be called `dataChannel`. Note that there are some rows that are unclassified according to the six channels of interest and those are indicated by `other`.   
  
  Once the data column is created, we can easily subset the data using the `filter` function to create a new data set for each data channel. We removed the original `data_channel_is_*` columns as well as two non-predictive columns `url` and `timedelta`.

```{r dataImport}

# reading in the data set
rawData <- read_csv("../OnlineNewsPopularity.csv")

# creating new variable to have more comprehensive names for data channels.
rawDataChannel <- rawData %>%
  mutate(dataChannel = ifelse(data_channel_is_lifestyle == 1, "lifestyle", 
                              ifelse(data_channel_is_entertainment == 1, "entertainment", 
                              ifelse(data_channel_is_bus == 1, "bus", 
                              ifelse(data_channel_is_socmed == 1, "socmed", 
                              ifelse(data_channel_is_tech == 1, "tech", 
                              ifelse(data_channel_is_world == 1, "world", 
                                     "other"))))))) %>%
  select(-data_channel_is_lifestyle, -data_channel_is_entertainment, 
         -data_channel_is_bus, -data_channel_is_socmed, -data_channel_is_tech,
         -data_channel_is_world, -url, -timedelta)

# assigning channel data to R objects.
lifestyleData <- rawDataChannel %>%
  filter(dataChannel == "lifestyle")

entertainmentData <- rawDataChannel %>%
  filter(dataChannel == "entertainment")

busData <- rawDataChannel %>%
  filter(dataChannel == "bus")

socmedData <- rawDataChannel %>%
  filter(dataChannel == "socmed")

techData <- rawDataChannel %>%
  filter(dataChannel == "tech")

worldData <- rawDataChannel %>%
  filter(dataChannel == "world")

```


## Select Data for Appropriate Data Channel

To select the appropriate data channel based on the `params$channel`, we created a function `selectData` which would return the appropriate data set and assign it to the data set `activeData`. This will be the file we will use for the remainder of the report.   
  
  To select the appropriate data channel based on the `params$channel`, we created a function `selectData` which would return the appropriate data set and assign it to the data set `activeData`. This will be the file we will use for the remainder of the report.

```{r selectData}

# function to assign automated calls for the different data channels
selectData <- function(dataChannel) { 
  if (dataChannel == "lifestyle"){
    return(lifestyleData)
  }
  if (dataChannel == "entertainment"){
    return(entertainmentData)
  }
  if (dataChannel == "bus"){
    return(busData)
  }
  if (dataChannel == "socmed"){
    return(socmedData)
  }
  if (dataChannel == "tech"){
    return(techData)
  }
  if (dataChannel == "world"){
    return(worldData)
  }
}

# activating corresponding data set.
dataChannelSelect <- params$channel

activeData <- selectData(dataChannelSelect)

```


## Summarizations for data channel *`r params$channel`*

In this section, we will perform EDA for the data channel `r params$channel`

### Data Split

This section splits the data set into training and test sets for the proportion of 70/30. The data summarization will be conducted on the training set. To split the data, the function `createDataPartition()`, from `caret` package, was used with the argument `p=0.7` to represent 80% of the data should be in the split. The function `set.seed(555)` was used to fix the random seed. The code below shows the creation of training and test sets.

```{r splitData}

set.seed(555)

trainIndex <- createDataPartition(activeData$shares, p = 0.7, list = FALSE)

activeTrain <- activeData[trainIndex, ]

activeTest <- activeData[-trainIndex, ]

```


### Data manipulation for statistics

A new object is created in this section aiming to summarize publications during weekdays and weekends and create factor levels for them to match with `shares` variable. The functions `ifelse()` was used to vectorize the IF-ELSE statements associated to `mutate()` which took care of attaching the new variable to the data set. The function `factor()` was used to explicitly coerce the days of week into levels of the newly created categorical variable "Day".

```{r statsData}

# IF-ELSE statements
statsData <- activeTrain %>%
  mutate(Day = ifelse(weekday_is_monday == 1, "Monday", 
                      ifelse(weekday_is_tuesday == 1, "Tuesday", 
                      ifelse(weekday_is_wednesday == 1, "Wednesday", 
                      ifelse(weekday_is_thursday == 1, "Thursday", 
                      ifelse(weekday_is_friday == 1, "Friday", 
                      ifelse(weekday_is_saturday == 1, "Saturday", 
                      ifelse(weekday_is_sunday == 1, "Sunday",
                             "missingdata")))))))) %>%
  mutate(Weekend = ifelse(is_weekend == 1, "Yes", "No"))

# Assigning factor levels
statsData$Day <- factor(statsData$Day, 
                levels = c("Monday", "Tuesday", "Wednesday", "Thursday", 
                           "Friday", "Saturday", "Sunday"))

```

### Belknap - Summary Statistics

The following table gives us information about the summary statistics for the number of shares for articles in the data channel `r params$channel`. The `summary()` function was used to extract these metrics.

```{r summaryStats}

summary(activeTrain$shares)

```

The following table gives us information about the average, median, and standard deviation for the number of shares based on whether the post was made on a weekend or a weekday. The variable "weekend" was grouped, via `grouped_by()`, and for each level the sum, average, median, and standard deviation of shares were calculated via `sum()`, `mean()`, `meadian()`, `sd()`, and `summarise()` functions. The summary table is shown below.

```{r table1}

statsData %>% 
  group_by(Weekend) %>%
  summarise(sumShares = sum(shares), avgShares = mean(shares), medShares = median(shares), sdShares = sd(shares))

```

Likewise, this table gives us information about the number of shares by the day of the week. The same functions were used here, by applied to levels of variable "Day". Also, the quantities maximum `max()` and minimum `min()` number of shares by levels of "Day" were calculated.

```{r table2}

statsData %>% 
  group_by(Day) %>%
  arrange(Day) %>%
  summarise(sumShares = sum(shares), avgShares = mean(shares), medShares = median(shares), sdShares = sd(shares), maxShares = max(shares),
            minShares = min(shares))

```


### Monti - Summary Statistics


First, we will analyse the frequency of occurrence of publications on each day of the week. The one-way contingency table below presents those frequencies.

```{r}

table(statsData$Day)

```

A second discrete analysis performed here is the two-way contingency table related to the discretization of the response variable if we divided `shares` into two categories. The function `cut()` was used for the end. In this case, we count the frequency of the number of publications in days of week with the two levels of response variable. These levels represent smaller number of shares (on the left) and larger number of shares (on the right). The table below shows this counting.

```{r}

table(statsData$Day, cut(statsData$shares, breaks = 2))

```

An important EDA analysis for regression tasks is the calculation of correlation matrix. The function `cor()` is used in this section to return the top 10 most correlated potential predictor variables with the response variable `shares`. The code below presents the process of obtaining these variables and their respective correlations with the response variable. 

```{r correl}

var_sel = select(activeTrain,starts_with("LDA_"), average_token_length,
         is_weekend, n_tokens_content, n_non_stop_unique_tokens, num_hrefs,
         num_self_hrefs, num_videos, average_token_length, kw_avg_min, 
         kw_avg_max, kw_avg_avg, is_weekend)

# correlation matrix
correlation = cor(activeTrain$shares, var_sel)

# sorting the highest correlations
p = sort(abs(correlation), decreasing = T)

# getting column ID
var_id = which(abs(correlation) %in% p[1:10])

# collecting variable names
var_cor = colnames(correlation)[var_id]

#combining names with correlations
tbcor = cbind(var_cor, correlation[var_id])

# converting to tibble
tbcor = as_tibble(tbcor)

# updating column names
colnames(tbcor)=c("Variables","Correlation")

# rounding the digits
tbcor$Correlation = round(as.numeric(tbcor$Correlation),3)

# nice printing with kable
kable(tbcor, caption = "Top 10 Response Correlated Variables")

```

The variables that present most correlation with the response variable `shares` are `r var_cor`. These variables will be studied in more depth via PCA to understand the orientation of the most variable potential predictors. The code below presents the PCA analysis as part of the EDA.

```{r PCA}

id = which(colnames(activeTrain) %in% var_cor)

# PCA
PC = prcomp(activeTrain[,id], center = TRUE, scale = TRUE)

pc_directions=as.data.frame(PC$rotation)

kable(pc_directions, caption="PCs for EDA", digits = 3)

```


### Monti - Graphs (3)

The plot below presents histograms, scatter plots, and correlations in a bivariate structure of the top 5 variables chosen the correlation analysis. Notice the shape of the distributions and the values of the correlations for the first column, which the one related to the response variable `shares`.

```{r correlplot}

# bivariate correlation plot
cor_data <- select(activeTrain,shares,var_id[1:5])
chart.Correlation(cor_data, histogram=TRUE)

```

The biplot below presents the PC1 and PC2 from the PCA analysis. The function `ggplot()` was used to create the plot and the segments created via `geom_segment()` were rescaled so that we could better see the variable names. The most variation in the data is contained in the PC1, hence, the most important variables in the data approximately are oriented towards the axis of PC1 and, therefore, may be good predictors for the `shares` response. Likewise, for PC2, which contains the second most variability in the data set, the variables that are oriented approximately towards the axis of PC2 are the second most important variables.

```{r biplot}

pc_df<-data.frame(PC$x)
# plotting PC1 and PC2 for the top 5 variables
# biplot(PC, cex = 1)
ggplot(pc_directions)+
  geom_point(data = pc_df, mapping = aes(x=PC1, y=PC2))+
  geom_segment(aes(x = 0, y = 0, yend = 50 * PC2, xend = 50 * PC1))+
  geom_label(mapping = aes(, x = 51 * PC1, y = 51 * PC2, label = row.names(pc_directions)))

```

The scatter plots below show the different levels of the variables related to the LDA metrics, from 0 to 4, and graphs the relationship with the response variable `shares`. The function `ggplot()` is used to create the plot frame and `geom_point()`, `geom_smooth`, and `facert_wrap()` function are used to plot the scatter points, the smooth GAM (Generalized Additive Models) lines, and split the data by LDA type, respectively. It is possible to see the behavior of the response variable in relation to each LDA types.

```{r LDAplot}

LDA.dat = activeTrain %>%
  select(shares, starts_with("LDA")) %>%
  pivot_longer(cols = LDA_00:LDA_04, names_to = "LDA", values_to = "values")

# relationship between shares and LDA levels (facet_wrap+smooth)
ggplot(LDA.dat, aes(y = shares, x = values))+
  geom_point() + geom_smooth(method = "loess")+ facet_wrap(~LDA)+
labs(x = "LDA Values", y = "Shares", title = "Shares by LDA Types")


```

The scatter plots below show the different types of the variables related to the Keyword metrics and graphs the relationship with the response variable `shares`. The function `ggplot()` is used to create the plot frame and `geom_point()`, `geom_smooth`, and `facert_wrap()` function are used to plot the scatter points, the smooth GAM (Generalized Additive Models) lines, and split the data by keyword type, respectively. It is possible to see the behavior of the response variable in relation to each of the 3 keyword metric types.

```{r keywordplot}

# relationship between shares and keyword metrics
kw.dat = activeTrain %>%
  select(shares, kw_avg_max, kw_avg_avg, kw_avg_min) %>%
  pivot_longer(cols = 2:4, names_to = "keyword", values_to = "values")

# relationship between shares and keyword metrics types (facet_wrap+smooth)
ggplot(kw.dat, aes(y = shares, x = values))+
  geom_point() + geom_smooth(method = "loess")+ facet_wrap(~keyword)+
labs(x = "Keyword Metric Values", y = "Shares", title = "Shares by Keyword Metric Types")

```

Finally, the scatter plots below show the different types of the variables related to the Content metrics and graphs the relationship with the response variable `shares`. The function `ggplot()` is used to create the plot frame and `geom_point()`, `geom_smooth`, and `facert_wrap()` function are used to plot the scatter points, the smooth GAM (Generalized Additive Models) lines, and split the data by content type, respectively. It is possible to see the behavior of the response variable in relation to each of the 4 content metric types.

```{r Contentplot}

# relationship between shares and content metrics (facet_wrap+smooth)
cont.dat = activeTrain %>%
  select(shares, num_videos, n_tokens_content, n_non_stop_unique_tokens,
         num_hrefs, num_self_hrefs, average_token_length) %>%
  pivot_longer(cols = 2:7, names_to = "content", values_to = "values")

# relationship between shares and content metrics types (facet_wrap+smooth)
ggplot(cont.dat, aes(y = shares, x = values))+
  geom_point() + geom_smooth(method = "loess")+ facet_wrap(~content)+
labs(x = "Content Metric Values", y = "Shares", title = "Shares by Content Metric Types")

```


### Belknap - Graphs (3)

The following graph shows the number of shares compared to the number of words in the title. The output is colored by the day of the week. 

```{r titlewordcountGraph}

titlewordcountGraph <- ggplot(statsData, aes(x = n_tokens_title, y = shares))
titlewordcountGraph + geom_point(aes(color = Day)) + 
  ggtitle("Number of Shares vs. Number of Words in Title") +
  ylab("Number of Shares") +
  xlab("Number of Words in Title")

```

The following plot shows the number of shares by the rate of positive words in the article. A positive trend would indicate that articles with more positive words are shared more often than articles with negative words.

```{r positivewordrateGraph}

positivewordrateGraph <- ggplot(statsData, aes(x = rate_positive_words, y = shares))
positivewordrateGraph + geom_point(aes(color = Day)) + 
  ggtitle("Number of Shares vs. Rate of Positive Words") +
  ylab("Number of Shares") +
  xlab("Rate of Positive Words") 

```

The following plot shows the total number of shares as related to the parameter title subjectivity. A positive trend would indicate that articles are shared more often when the title is subjective. A negative trend would indicate that articles are shared more often when the title is less subjective.

```{r titleSubjectivityGraph, eval=FALSE}

titleSubjectivityGraph <- ggplot(statsData, aes(x = title_subjectivity, y = shares))
titleSubjectivityGraph + geom_point(aes(color = n_tokens_title)) + 
  ggtitle("Number of Shares vs. Title Subjectivity") +
  ylab("Number of Shares") +
  xlab("Title Subjectivity") + 
  labs(color = "Word Count in Title")

```

### Subsetting Variables for Modeling

The variables selected below are those described in the introduction of this study and will be used in the modeling section. The function `select()` was used to subset the corresponding variables from the training and test sets and two new objects are created specially for the modeling section, `dfTrain` and `dfTest`.

```{r choseVars}

dfTrain = activeTrain %>%
  select(shares, starts_with("LDA_"), average_token_length,
         is_weekend, n_tokens_content, n_non_stop_unique_tokens, num_hrefs,
         num_self_hrefs, num_videos, average_token_length, kw_avg_min, 
         kw_avg_max, kw_avg_avg, is_weekend)

dfTest = activeTest %>%
  select(shares, starts_with("LDA_"), average_token_length,
         is_weekend, n_tokens_content, n_non_stop_unique_tokens, num_hrefs,
         num_self_hrefs, num_videos, average_token_length, kw_avg_min, 
         kw_avg_max, kw_avg_avg, is_weekend)

```


## Modeling

In this section, we will perform regression for prediction purposes for the data channel `r params$channel`. All models were fitted using 5-fold Cross-Validation via `train()` function from `caret` package.

### Belknap - Linear Regression Model Explanation

(add some thoughts here)

### Monti - Linear Regression Model (LASSO Regression)

The linear regression chosen for this next model is based on penalized regression via LASSO method. This method has a particular advantage of having a triangular shape of parameters search space so that it allows the estimated coefficients to be zero. Hence, LASSO regression is also a variable selection method. In this application, we will test the prediction capability of LASSO regression only. It was tested a sequence of values for the Regularization Parameter (lambda), a tuning parameter, from 0 to 10 by 1 via `seq(0,10,1)` assigned to the `tuneGrid = `argument in the `train()` function from `caret` package. The code below presents the estimated coefficients for the best hyperparameter.

```{r LASSOtrain}

LASSO = train(shares~., data = dfTrain,
              method="glmnet",
              preProcess = c("center","scale"),
              tuneGrid = expand.grid(alpha = 1, lambda = seq(0,10,1)),
              trControl = trainControl(method="CV",number=5))

coef(LASSO$finalModel, LASSO$bestTune$lambda)

```

The best lambda for this model is `r LASSO$bestTune$lambda` and this value can be seen in the table below which summarizes all the metrics for the 5-fold cross-validation.

```{r LASSOresultTrain}

lasso_out = data.frame(LASSO$results)

kable(lasso_out, caption = "Output Training Metrics for LASSO",
      digits = 3)

```

The plot below shows the RMSE by Regularization Parameter (lambda). It is easy to see that RMSE is minimized when $\lambda$ = `r LASSO$bestTune$lambda`.

```{r LASSOplot}

plot(LASSO)

```

The validation step for LASSO regression is applied on the test set after predicting the response variable for unseen data (test set). By using `predict()` and `postResample()` functions, the metrics RMSE (Root Means Squared Error), R2 (Coefficient of Determination), and MAE (Mean Absolute Error) are calculated and displayed below.

```{r LASSOtest}

metric_LASSO = postResample(pred = predict(LASSO, newdata = dfTest),
                            obs = dfTest$shares)

metric_LASSO

```

### Belknap - Linear Regression Model

(add some thoughts here)

```{r linearTrain}

lmFit = train(shares~., data = dfTrain,
              method="lm",
              preProcess = c("center","scale"),
              trControl = trainControl(method="CV",number=5))

summary(lmFit)

```

(add some thoughts here)

```{r linearTrainResult}

lm_out = data.frame(lmFit$results)

kable(lm_out, caption = "Output Training Metrics for Linear Regression",
      digits = 3)

```


```{r lineartest}

metric_lm = postResample(pred = predict(lmFit, newdata = dfTest), 
                         obs = dfTest$shares)

metric_lm

```


### Belknap - Random Forest Model & Explanation

(add some thoughts here about RF)

```{r RFtrain}

train.control = trainControl(method = "cv", number = 5)

rfFit <- train(shares~.,
               data = dfTrain,
               method = "rf",
               trControl = train.control,
               preProcess = c("center","scale"),
               tuneGrid = data.frame(mtry = 1:5))

rfFit$bestTune$mtry

```

(add some thoughts here)

```{r RFplot}

plot(rfFit)

```

(add some thoughts here)

```{r RFresults}

rf_out = data.frame(rfFit$results)

kable(rf_out, caption = "Output Training Metrics for Random Forest",
      digits = 3)

```

(add some thoughts here)

```{r RFTest}

RF_pred <- predict(rfFit, newdata = activeTest)

metric_rf = postResample(RF_pred, activeTest$shares)

metric_rf

```


### Monti - Boosted Tree Model & Explanation

(add some thoughts here)

```{r boostingTrain}

tunG = expand.grid(n.trees = seq(25,200,25),
                      interaction.depth = 1:4,
                      shrinkage = 0.1,
                      n.minobsinnode = 10)

gbmFit <- train(shares~.,
               data = dfTrain,
               method = "gbm",
               preProcess = c("center","scale"),
               trControl = train.control,
               tuneGrid = tunG,
               verbose = FALSE
               )


gbmFit$bestTune$n.trees

gbmFit$bestTune$interaction.depth

```

The best n.trees and interaction.depth parameters for this model are `r gbmFit$bestTune$n.trees` and `r gbmFit$bestTune$interaction.depth`, respectively. These values can be seen in the table below, which summarizes all the metrics for the 5-fold cross-validation. It is easy to see that these values minimize the RMSE.

```{r boostingResults}

gbm_out = data.frame(gbmFit$results)

kable(gbm_out, caption = "Output Training Metrics for Boosting",
      digits = 3)

```

The plot below shows the RMSE by Number of Boosting Iterations and display Max Tree Depth lines for the 5-fold CV performed. It is easy to see that RMSE is minimized when n.trees = `r gbmFit$bestTune$n.trees` and interaction.depth = `r gbmFit$bestTune$interaction.depth`.

```{r boostingPlot}

plot(gbmFit)

```

The validation step for Boosting is applied on the test set after predicting the response variable for unseen data (test set). By using `predict()` and `postResample()` functions, the metrics RMSE (Root Means Squared Error), R2 (Coefficient of Determination), and MAE (Mean Absolute Error) are calculated and displayed below.

```{r boostingTest}

gbm_pred <- predict(gbmFit, newdata = activeTest)

metric_boosting = postResample(gbm_pred, activeTest$shares)

metric_boosting

```


## Comparison & Conclusion - Monti

For the overall comparison among all 4 created models in previous sections, the test set was used for predictions and some quality of fit metrics were calculated based on these prediction on unseen data. The code below shows the function that returns the name of the best model based on RMSE values estimated on the test set. The code below displays the table comparing all 4 models.

```{r comparison}

bestMethod = function(x){
  
  bestm = which.min(lapply(1:length(x), function(i) x[[i]][1]))
  
  out = switch(bestm,
                "Random Forest",
                "Boosting",
               "LASSO Regression",
               "Linear Regression")
  
  return(out)
  
}

tb = data.frame(RF = metric_rf, Boosting = metric_boosting,
                LASSO = metric_LASSO, Linear = metric_lm)

kable(tb, caption = "Accuracy Metric by Ensemble Method on Test Set",
      digits = 3)

```


After comparing all the 4 models fit throughout this analysis, the best model was chosen based on the RMSE value, such that the model with minimum RMSE is the "winner". Therefore, the best model is `r bestMethod(tb)` based on RMSE metric. The RMSE, R2 (coefficient of Determination), and MAE metrics for all 4 models can be seen in the table above.


